---
title: "Lesson 19: Inference for Independence of Categorical Data"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: false
---

<script type = "text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
</script>

<!--
<div style = "float:right;width = 40%;">
<br/>
<div style = "padding-left:10%;">**Optional Lesson Video**</div>
<iframe width = "90%" align = "right" src = "https://www.youtube.com/embed/videoseries?list = PLaZryQtbPQC9acEfJHNYCs9N93R2CJ8zB" frameborder = "1" allow = "autoplay; encrypted-media" allowfullscreen></iframe>
</div>
-->

## Lesson Outcomes

<a href = "javascript:showhide('oc')"><span style = "font-size:8pt;">Show/Hide Outcomes</span></a>
<div id = "oc" style = "display:none;">
By the end of this lesson, you should be able to do the following.

* Perform a hypothesis test for the Chi-square test for independence using the following steps:
  + State the null and alternative hypotheses
  + State the level of significance ($\alpha$)
  + Create numerical and graphical summaries of the data
  + Calculate the test-statistic, degrees of freedom and P-value of the test using software
  + Assess statistical significance by comparing the P-value to the level of significance ($\alpha$)
  + State the appropriate conclusion for the hypothesis test
  + Check the requirements for the hypothesis test

</div>
<br>



<div style = "clear:both;"></div>
<br/>

## The Chi-squared Test of Independence

The chi-squared test of independence allows us to determine if two different categorical variables are related to each other. Here are some questions that could be answered with a chi-squared test of independence.

* Does the education level obtained by your father impact the education level you will obtain in life? (Here education level could be defined to be one of: High School, Associates Degree, Bachelor's Degree, or Graduate Degree. The two variables we would be comparing would be the child's education level and their father's education level.)
* Does the marital status of a person (single, married, widowed, or divorced/separated) relate to the overall happiness of that person (very happy, happy, not too happy)?
* Does a student's major (say Business, Nursing, Biology, or Math) have any impact on their ability to get a good paying job (say above \$50,000 per year versus below \$50,000) after graduating from college?

In each of these examples, there were two sets of categorical data that were being compared to see if there was any relation between those variables. An important word that will be used in this lesson to talk about variables being related or not is the word *independent*. Things that **are independent** of each other have absolutely no relation to each other. On the other hand, when things are related to each other, that means that they are **not independent of each other**. Keep this in mind as you read this lesson. Independent means not related, and *not independent* means *related*.



<br>


<img src = "./Images/StepsAll.png">

### Case Study: Chiropractic Care


<div style = "float:right;padding:10px;">
<img src = "./Images/chiropracticCare.jpeg" width = "400px;">
</div>


<!-- (Reasons for wording it this way will be given after you've been through the entire hypothesis test.) -->

<!--
To prepare for the hypothesis test, recall that our hypothesis tests have assessed how different two or more things are. For example, in the 2-sample $t$-test, we compare two population means by seeing how different they are. In a test of one proportion, we measure the difference between a sample proportion and a population proportion. Each of these tests requires some information about at least one parameter. That information is given in the null hypothesis of the test, which we've always been able to write with one or more " = " signs, because we've always used numeric parameters.

Unfortunately, we have no parameter we can use to measure the independence of the variables "location" and "motivation". Nevertheless, we have to be able to calculate a $P$-value, and the only approach we have so far is to measure differences between things. "Difference" means "subtraction" in Statistics, and if the measurements are categorical, we cannot subtract them. Therefore, as in the other lessons of this unit, we will count occurrences of each value of each variable to get some numbers. (The example that follows will show how this is handled.) The numbers we get will be called "counts" or "observed counts".
-->


<img src = "./Images/Step1.png">

A [study was conducted](https://drcharlesblum.com/About%20Us/BlumJCCASep2008.pdf) to determine why patients in Australia, Europe, and the U.S.A. seek chiropractic care<!--<cite>Blum08</cite>-->.  Patients were classified based on their location (Australia, Europe, or the U.S.A.) and their motivation for seeking treatment.  Using descriptions developed by Green and Krueter, patients were asked which of the five reasons led them to seek chiropractic care <!--<cite>Blum08,Green99</cite>-->:

- *Wellness*: defined as optimizing health among the self-identified healthy
- *Prevention*: defined as preventing illness among the self-identified healthy
- *At risk*: defined as preventing illness among the currently healthy who are at heightened risk to develop a specific condition
- *Sick role*: defined as getting well among those self-perceived as ill with an emphasis on therapist-directed treatment
- *Self care*: defined as getting well among those self-perceived as ill favoring the use of self vs. therapist directed strategies

The research question was whether people's motivation for seeking chiropractic care was related to their location: Europe, Australia, or the United States or not. 

People seek chiropractic care for different reasons. We may want to know if those reasons are different for Europeans than for Americans or Australians. Do reasons for seeking chiropractic care depend on the location in which one lives?

If the reasons people seek chiropractic care are the same in each of the regions, we can say "Motivations for seeking chiropractic care and one's location are independent." (The statistical meaning of "independent" is too technical to give here. However, for now, you can think of it as knowing the value of one variable gives you no information about the value of the other variable.) This is the null hypothesis of the test we will conduct. We can write the alternative hypothesis as the opposite: "Motivations for seeking chiropractic care and one's location are not independent." 

The null and alternative hypotheses for this chi-squared test of independence are:
$$
\begin{array}{rl}
H_0\colon & \text{The location and the motivation for seeking treatment are independent} \\
H_a\colon & \text{The location and the motivation for seeking treatment are not independent} \\
\end{array}
$$

The data we will examine is the number of people observed in each region seeking chiropractic treatment for specific reasons.
For simplicity, we arrange counts of the number of observations of each motivation in each region in a table called a "contingency table." One variable is chosen as the "row variable," so called because its values are the row headers for the table. The other variable is called the "column variable," because its values are the column headers.

Note: When speaking of the hypotheses in the absence of a context, we can write them in the form
$$
\begin{array}{rl}
H_0\colon & \text{The row variable and the column variable are independent} \\
H_a\colon & \text{The row variable and the column variable are not independent} \\
\end{array}
$$
When writing these in the context of a given situation, replace "row variable" and "column variable" by a more descriptive reference.

We will use a level of significance of $\alpha = 0.05$.

<br>

<img src = "./Images/Step2.png">

The researchers reported on their data collection procedures with the following statement. "Using an international convenience sample of Sacro-Occipital Technique (SOT) practitioners, 1316
consecutive patients attending 27 different chiropractic
clinics in the USA, Europe and Australia completed a
one-page survey on intake to assess reason for seeking
care. A forced choice response was obtained
characterizing the patient’s reason for seeking
chiropractic care."

<br/>

<img src = "./Images/Step3.png">


```{r, include = FALSE, eval = FALSE}
# # The following code was used to generate the untallied data:
# library(tidyverse)
# library(readxl)
# library(writexl)
# url <- "https://byuistats.github.io/M221R/data/chiropractic_care_table.xlsx"
# destfile <- "chiropractic_care_table.xlsx"
# curl::curl_download(url, destfile)
# chiropractic_care_table <- read_excel(destfile)
# View(chiropractic_care_table)
# 
# chiropractic_care_table %>%
#   select(-Source) %>% 
#   pivot_longer(cols = 2:ncol(.)) %>% 
#   uncount(value) %>% 
#   rename(location = "Location \\ Motivation", motivation = name) %>%
#   mutate(rand = runif(n())) %>%
#   arrange(location, rand) %>%
#   select(location, motivation) %>% 
#   write_xlsx("chiropractic_care_raw.xlsx")
```

```{r, echo = FALSE, message = FALSE}
library(readxl)
library(knitr)
library(tidyverse)
library(kableExtra)
dots <- data.frame(location = ":", motivation = ":")
chiropractic_care_raw <- read_excel("./data/chiropractic_care_raw.xlsx") %>% data.frame()
```

The data from the study are given in the file [chiropractic_care_raw.xlsx](./data/chiropractic_care_raw.xlsx). (Review of [reading in data](RHelp.html#reading-in-the-data).)
Each row represents the responses of one person. 

You can use the `head()` function to look at the first few rows of the data.
```{r, comment = NA}
head(chiropractic_care_raw)
```

You can also view the data using the command `View(chiropractic_care_raw)`.
Here is a brief excerpt of the data:
```{r, echo = FALSE}
chiropractic_head <- chiropractic_care_raw %>%
  head(6)
chiropractic_mid <- chiropractic_care_raw %>%
  filter(location == "Europe") %>%
  head(4)
chiropractic_tail <- chiropractic_care_raw %>%
  tail(4)
bind_rows(chiropractic_head, dots, chiropractic_mid, dots, chiropractic_tail) %>%
  kable() %>%
  kable_styling(full_width = F)
```



**Motivation for Choosing to Seek Chiropractic Care** 


### R Instructions for Clustered Bar Charts

<div class = "SoftwareHeading">R Instructions</div>
<div class = "Software">

First, read in the data from the file [chiropractic_care_raw.xlsx](./data/chiropractic_care_raw.xlsx). (Review of [reading in data](RHelp.html#reading-in-the-data).)

It is hard to visualize these data. We summarize them into a table of counts, which can be called a *contingency table*. <!--<cite>Blum08</cite>-->

After reading in the data, we can organize it into a contingency table. The values in a contingency table represent the counts of the reasons individuals from each location sought chiropractic care.

```{r}
contingency_table <- table(chiropractic_care_raw)
contingency_table
```

<!--
```{r, echo = FALSE, message = FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
chiropractic_care_raw %>% 
  table() %>% 
  kable() %>% 
  kable_styling(full_width = F) %>% 
  column_spec(1, bold = TRUE)
```
-->

We can transpose (or swap the rows and columns of) this table using the command `t(...)`.
```{r}
t(contingency_table)
```

Notice that the values in the first column of the contingency table are in the first row of its transpose. When a table is transposed, the rows and columns essentially trade roles.

While a table of counts (a contingency table) is helpful in describing the relationship between two categorical variables like *location* and *motivation*,  a clustered bar chart makes it much easier to visually compare and contrast the observed counts.
We can use the contingency table to create a clustered bar chart using the `barplot( )` command. We will include the following optional commands:

* `beside = TRUE` puts the bars side-by-side. Not using `beside = TRUE` results in a stacked bar chart.

* `legend.text = TRUE` adds a legend to the graph. Sometimes the legend covers the bars, so it is useful in that case to use `ylim = c(0, some number)` so that the legend is moved off of the bars.

```{r}
barplot(contingency_table, beside = TRUE, legend = TRUE)
```

It is usually helpful to also examine a bar chart of the transpose of the data.

```{r}
barplot(t(contingency_table), beside = TRUE, legend = TRUE)
```

Either plot can be used, because they present identical information in different ways. You will probably prefer one of these plots over the other, so use the one you feel best describes the data. 

In this case, it makes sense to group the reasons people seek chiropractic care together for each country. The second boxplot seems to be more effective as presenting the information in this case.

Notice that the legend covers part of the bars in this plot. If we increase the length of the y-axis to 450 (instead of the default value), the legend no longer covers the bars. This is accomplished with the argument `ylim = c(0, 450)`, which sets the limits of the y-axis to go from 0 to 450. We can also color the plot by providing enough colors for each of the legend entries. We will use the first plot, which has five legend entries. So, we use the command: `col = c("tan3", "brown3", "darkorchid", "steelblue3", "darkolivegreen4")`.

```{r}
barplot(t(contingency_table), beside = TRUE, legend = TRUE,
        ylim = c(0, 450),  # extend the y-axis from 0 to 450
        col = c("tan3", "brown3", "darkorchid", "steelblue3", "darkolivegreen4"))
```

<br/>
</div>

<br>
<br>

The predominant motivation that people choose to seek chiropractic care in any region is for *Self Care*. The second-most common reason differs across the three regions. For Australia, it is *At Risk*. For Europe, it is *Sick Role*, and for the United States, it is *Wellness*. 

<br/>

<img src = "./Images/Step4.png">

Our objective is to determine if the reasons people seek chiropractic care is proportionally the same for each of the three locations. In other words, we want to know if people's motivation for seeking chiropractic care is independent of their location. 

The barcharts for the three regions do not look the same. This suggests that the motivation to seek chiropractic care is not independent of the region. 
However, before drawing a conclusion it is important to conduct the chi-squared test of independence. The $P$-value from this test will help us know if there are statistically significant differences in the ratios across the groups. 
If we fail to reject the null hypothesis, then *motivation* and *location* are independent (i.e., not related). So any patterns we are seeing in the bar plot can be attributed to random chance. However, if we reject the null hypothesis then we can claim that *motivation* and *location* are **not independent** (meaning the proportions of the reasons people seek care is different in the three locations). We can then draw our final conclusions regarding the patterns observed in the side-by-side bar plots.



### R Instructions for Performing a Chi-squared Test of Independence

<div class = "SoftwareHeading">R Instructions</div>
<div class = "Software">

To perform a chi-squared test of independence in R, use the `chisq.test( )` function. Using this function requires that you have previously made a table of counts. (Review the [R Instructions for Clustered Bar Charts](Lesson19.html#r-instructions-for-clustered-bar-charts) to review how to create a table of counts in R.)

```{r, comment = NA}
chisq_results <- chisq.test(contingency_table)
chisq_results
```

The "X-squared" value of 49.743 is the test statistic of the test. Technically, it should be typeset as $\chi^2 = 49.743$. The text "X-squared" is used to represent the symbol ($\chi^2$), which is the Greek letter chi ($\chi$) squared. (Note that $\chi$ is a Greek letter, and its name is pronounced "/ki/", as in "kite". The symbol $\chi^2$ should be pronounced "ki-squared.") The degrees of freedom (**df**) and the $P$-value are also given in this output. 
<!--
The value of the degrees of freedom is df = `r chisq_results$parameter` and the $P$-value is $4.58 \times 10^{-8} = 0.000~000~045~8$.
-->

</div>

<br/>


In this hypothesis test, R compares the counts of how many times things actually occurred compared to the counts we should *expect* to see, *if the null hypothesis is true.* We call the counts we would expect to see "expected counts." The software will then subtract the observed counts from the expected counts and combine these differences to create a test statistic that is used to get the $P$-value. That number is called the $\chi^2$ test statistic. 

In this course, we use software to calculate the $\chi^2$ test statistic, but it is important to recognize that the $\chi^2$ test statistic compares the observed counts to the expected counts---that is, to the counts we should expect to get if the null hypothesis is true. The larger the $\chi^2$ test statistic is, the smaller the $P$-value will be. If the $\chi^2$ test statistic is large enough that the $P$-value is less than $\alpha$, we will conclude that the observed counts and expected counts are too different for the null hypothesis to be plausible, and will therefore reject $H_0$. Otherwise, we will fail to reject $H_0$, as with other hypothesis tests.



Different $\chi^2$ distributions are distinguished by the number of degrees of freedom, which are determined by the number of rows and columns in the table: 

$$df = (\text{number of rows }-1)(\text{number of columns }-1)$$

One way to compute the degrees of freedom is to cross out the last row and the last column of the contingency table and to count the number of cells that remain. 
Note that the number of degrees of freedom does not depend on the number of subjects in the study.


Recall that the hypotheses of this test were:

$~~~~~H_0\colon$ Location and the motivation to visit a chiropractor are independent

$~~~~~H_a\colon$ Location and the motivation to visit a chiropractor are not independent

The test statistic is: $\chi^2 = 49.743$, with $df = 8$ and $P$-value $= 4.58 \times 10^{-8} = 0.000~000~045~8 < 0.05 = \alpha$. We reject the null hypothesis and conclude that location and the motivation to visit a chiropractor are not independent. This means that the motivation differs across the locations.


### Requirements

The following requirements must be met in order to conduct a $\chi^2$ test of independence:

1. The sample of data should be representative of the full population. This is most likely to happen when simple random sampling is used to obtain a sample from the population.

    * In this study, a convenience sample was used. So there could be some doubt as to whether these individuals are representative of all individuals in each country that seek chiropractic care.

2. Each expected count must be greater than or equal to 5.

    * In this study, all expected counts were greater than 5.
    
    * To check expected counts in R, use the code:
    
    ```{r, comment = NA}
    chisq_results$expected
    ```

    * As shown in the above table, the smallest expected count is `34.92857`, which is greater than or equal to 5. 
    
We conclude the assumptions of the chi-squared test for independence are satisfied.
    

<img src = "./Images/Step5.png">

People in Europe, Australia, and the United States evidently have different reasons for seeking chiropractic care. It appears that individuals in Europe are less likely to seek chiropractic care for self care and more likely to seek care in the sick role than individuals in the other regions. The Europeans are also less likely to visit a chiropractor for wellness reasons.

<br>

## Other considerations

### Swapping the Row and Column Variables

There is no general guideline for deciding which variable is the row variable and which variable is the column variable in a $\chi^2$ test of independence. To see why not, complete the questions that follow.

<div class = "QuestionsHeading">Answer the following questions:</div>
<div class = "Questions">
1. Re-do the $\chi^2$ test of independence for the chiropractic care data, but use the transpose of the contingency table.  Then compare the degrees of freedom, $\chi^2$ test statistic, and $P$-value of this test, with the degrees of freedom, $\chi^2$ test statistic, and $P$-value for the test conducted above, when "Location" was the row variable.

<a href = "javascript:showhide('Q1')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q1" style = "display:none;">
- In both tests, $df = 8$, $\chi^2 = 49.743$, and the $P$-value is $4.58 \times 10^{-8}  < 0.05 = \alpha$. They are the same for both tests.
</div>
<br>

2. What do you conclude about swapping the row and column variables in a $\chi^2$ test of independence?

<a href = "javascript:showhide('Q2')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q2" style = "display:none;">
- Swapping the row and column variables in a $\chi^2$ test of independence does not change the conclusion of the test. 
</div>
&nbsp;
</div>
<br>



### Why $H_a$ is Worded As It Is

Recall that in the chiropractic care example, the hypotheses for the $\chi^2$ test of independence were

$$
\begin{array}{1cl}
H_0: & \text{Location and the motivation to visit a chiropractor are independent} \\
H_a: & \text{Location and the motivation to visit a chiropractor are not independent}
\end{array}
$$

You may wonder why we don't write "$H_a\colon$ The motivation to visit a chiropractor depends on location." Well, couldn't we say just as easily that location depends on the motivation to visit a chiropractor? It may seem a little strange when phrased this way. Let's use the following exercises to look briefly at a somewhat less strange example, then return to this example.

<div class = "QuestionsHeading">Answer the following questions:</div>
<div class = "Questions">
3. Suppose you want to know whether a student's stress level and the degree to which they feel a need to succeed are independent. What should your hypotheses be?

<a href = "javascript:showhide('Q3')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q3" style = "display:none;">
- $H_0\colon$ Stress level and the need to succeed are independent
- $H_a\colon$ Stress level and the need to succeed are not independent
</div>
<br>

4. For their alternative hypothesis, a student erroneously writes "$H_a\colon$ Stress level depends on the need to succeed." If they reject $H_0$, what will they conclude?

<a href = "javascript:showhide('Q4')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q4" style = "display:none;">
- They will conclude that a student's stress level depends on their need to succeed.
</div>
<br>

5. Another student erroneously writes "$H_a\colon$ Need to succeed depends on stress levels." If they reject $H_0$, what will they conclude?

<a href = "javascript:showhide('Q5')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q5" style = "display:none;">
- They will conclude that the degree to which a student feels a need to succeed depends on their stress level.
</div>
<br>

6. Could it be that a student's need to succeed depends on their stress level? Could it be that their stress level depends on their need to succeed? How can the $\chi^2$ test of independence distinguish between these two possibilities?

<a href = "javascript:showhide('Q6')"><span style = "font-size:8pt;">Show/Hide Solution</span></a>
<div id = "Q6" style = "display:none;">
- Students that feel a more intense need to succeed than others might very well experience higher stress levels. Likewise, a student who feels their stress levels rising might subconsiously feel that success will decrease their stress level, which could result in their feeling an increased need to succeed.

- The $\chi^2$ test statistic and the $P$-value will be the same, whether we write down the correct alternative hypothesis or one of the two erroneous ones mentioned above. The $\chi^2$ test of independence is not capable of telling whether stress levels depend on the need to succeed or whether the need to succeed depends on the stress level.

- If swapping the row and column variables made a difference in the outcome of the hypothesis test, then the test might be able to tell which variable depends on the other. But swapping the row and column variables does not change the outcome of the test.

- The full truth is worse than this: It may be that one's need to succeed and one's stress level depend on each other. It may be that neither depends on the other, but that both depend on something else, in a way that makes it look like they depend on each other. The $\chi^2$ test of independence is incapable of distinguishing among the many possible kinds of dependence.
</div>
&nbsp;
</div>
<br>

According to the exercises you just did, we are not justified in writing an alternative hypothesis that specifies which variable depends on which. Could we write "$H_a\colon$ Stress level and need to succeed are dependent"? After all, "independent" and "dependent" are opposites, aren't they? This may seem reasonable, but we have to be careful of the technical terms. Statisticians have gone to some trouble to carefully define "independent." They have not defined "dependent." (As suggested by the exercises you just did, dependence is complicated, perhaps too complicated to be able to be defined conveniently.) They use the phrase "not independent" as the opposite of "independent." So will we, writing "$H_a\colon$ Stress level and need to succeed are not independent."

Likewise, in the chiropractic care example, we can't say in the alternative hypothesis that location depends on motivation, nor that motivation depends on location, nor that each depends on the other, nor that both depend on something else, nor that location and motivation are dependent. Instead, we write "$H_a\colon$ The location and the motivation for seeking treatment are not independent," as statisticians do.



### No Confidence Intervals

We do not calculate confidence intervals when working with contingency tables. Think about it: With three rows and five columns in the table for the chiropractic care example, there are 15 proportions. It can be shown that if we compare these two at a time, there would be 105 pairs to compare. How could we possibly interpret a collection of 105 confidence intervals? Also, if our confidence level is 95%, we would expect that about 5 of our confidence intervals would not contain the true difference between proportions, but we wouldn't know which ones. Rather than take the risks this would cause, statisticians do not to calculate confidence intervals for contingency tables.

<!-- TO-DO LIST

## To-do list for the authors of this wiki
- Make sure all the vocabulary is defined.
- Add another example.
- Test of several proportions?
- Goodness-of-fit tests
- Lesson summary
- Add an optional section that shows how the $\chi^2$ statistic is calculated.

-->

<br>

## Summary

<div class = "SummaryHeading">Remember...</div>
<div class = "Summary">

- The **$\chi^2$ hypothesis test** is a test of independence between two variables. These variables are either associated or they are not. Therefore, the null and alternative hypotheses are the same for every test:
$$
\begin{array}{1cl}
H_0: & \text{The (first variable) and the (second variable) are independent} \\
H_a: & \text{The (first variable) and the (second variable) are not independent}
\end{array}
$$

- The **degrees of freedom ($df$)** for a $\chi^2$ test of independence are calculated using the formula $df = (\text{number of rows}-1)(\text{number of columns}-1)$

- In our hypothesis testing for $\chi^2$ we never conclude that two variables are *dependent*. Instead, we say that two variables are *not independent*.

- To create a clustered bar chart in R first requires using either `cbind( )` or `rbind( )` to create a table of observed counts, and then using 
`barplot(..., beside = TRUE, legend.text = TRUE)` to create the plot. Review this by [clicking here](Lesson19.html#r-instructions-for-clustered-bar-charts).

- To perform a chi-squared test of independence in R, use the `chisq.test( )` with your table of observed counts placed inside the function. Review this by [clicking here](Lesson19.html#r-instructions-for-performing-a chi-squared-test-of-independence).

- To check the requirements of a chi-squared test of independence, use the code `chisq_test_results$expected` where `chisq_test_results` is the results you saved from your chi-square test. Verify that all expected counts are greater than or equal to 5.

<br>
</div>
<br>

## Navigation

<center>
| **Previous Reading** | **This Reading** | **Next Reading** |
| :------------------: | :--------------: | :--------------: |
| [Lesson 18: <br> Inference for Two Proportions](Lesson18.html) | Lesson 19: <br> Inference for Independence of Categorical Data | [Lesson 20: <br> Review for Exam 3](Lesson20.html) |
</center>
